Finiteness in Cubical Type Theory

### 2020-06-30 sections/finiteness-predicates.tex
added a wrapfig

If that seems needlessly complex, don't worry: this definition is precisely
equivalent to the usual inductive one.
\begin{agdalisting}
  \ExecuteMetaData[agda/Data/List/Base.tex]{list-def}
\end{agdalisting}
And this isn't some kind of hand-waving equivalence, either: since we working in
HoTT, we can (and do) prove that the two types are equal, allowing us to use one
or the other depending on whichever is more convenient, and
\(\AgdaFunction{subst}\) in the other representation without loss of generality.
That said, defining lists as containers will reveal several interesting
connections and proofs about split enumerability and the other predicates, 
so for the remainder of the paper whenever we say \AgdaDatatype{List} we will
mean Equation~\ref{list-def}.

Before we get to those interesting proofs, though, there are some other things
that need defining.
\AgdaDatatype{Fin}, for instance: \(\AgdaDatatype{Fin}\;n\) is the type of
natural numbers smaller than \(n\).
Its definition relies on a type for disjoint union,
\(\AgdaDatatype{\ensuremath{\uplus}}\).
\begin{multicols}{2}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Data/Fin/Base.tex]{fin-def}
  \end{agdalisting} \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Introduction.tex]{disj-union}
  \end{agdalisting}%
  % It is also expressible with only \(\Sigma\):
  % \begin{agdalisting}
  %   \ExecuteMetaData[agda/Snippets/Introduction.tex]{sigma-disj-union}
  % \end{agdalisting}
  % Although the inductive type definition is slightly more ergonomic.
\end{multicols} \noindent
And containers themselves, of course.
Containers are a well-studied topic in dependent type theory, with a rich
theory: we won't dive in to that here.
\begin{definition}[Containers] \label{container-def}
  A container \cite{abbottContainersConstructingStrictly2005} is a pair
  \(S , P\) where \(S\) is a type, the elements of which are called
  the \emph{shapes} of the container, and \(P\) is a type family on \(S\), where
  the elements of \(P(s)\) are called the \emph{positions} of a container.
  We ``interpret'' a container into a functor defined like so:
  \begin{agdalisting} \label{container-interp}
    \ExecuteMetaData[agda/Container.tex]{container-interp}
  \end{agdalisting} \vspace{-\baselineskip}
\end{definition}

@@

As a slightly more complex example, consider the \(\AgdaDatatype{Fin}\) type
we've been using.
Remember that split enumerability is in fact the same as a split surjection from
\(\AgdaDatatype{Fin}\) (Lemma~\ref{split-enum-is-split-surj}): to show that
\(\AgdaDatatype{Fin}\) is split enumerable, then, we need only show that it has
a split surjection from itself.
We'll prove the following slightly more general statement:
\begin{agdalisting}
  \ExecuteMetaData[agda/Function/Surjective/Properties.tex]{split-surj-ident}
\end{agdalisting}
\subsection{Decidable Equality}
One thing that characterises all split enumerable types is that they are all
\emph{discrete}, i.e. they have decidable equality.  \todo{All multicol agda code should probably be vertically centred like this}
\begin{multicols}{2}  \null \vfill
  \begin{agdalisting}
    \ExecuteMetaData[agda/Relation/Nullary/Discrete/Base.tex]{discrete-def}
  \end{agdalisting} \vfill \null \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Dec.tex]{dec-def}
  \end{agdalisting}
\end{multicols} \vspace{-\baselineskip} \noindent
We will see later that this has implications for the space of types we're
dealing with, but for now it simply provides a useful function on split
enumerable types.

>>>>

\begin{wrapfigure}[5]{r}{.4\linewidth} \vspace{-1.6\baselineskip}
  \ExecuteMetaData[agda/Data/List/Base.tex]{list-def} \vspace{-\baselineskip}
  \caption{The Usual Inductive Definition}
\end{wrapfigure}

If that seems needlessly complex, don't worry: this definition is precisely
equivalent to the usual inductive one. %
% \begin{agdalisting}
%   \ExecuteMetaData[agda/Data/List/Base.tex]{list-def}
% \end{agdalisting}
And this isn't some kind of hand-waving equivalence, either: since we are
working in HoTT, we can (and do) prove that the two types are equal, allowing us
to use one or the other depending on whichever is more convenient, and
\(\AgdaFunction{subst}\) in the other representation without loss of generality.
That said, defining lists as containers will reveal several interesting
connections and proofs about split enumerability and the other predicates, 
so for the remainder of the paper whenever we say \AgdaDatatype{List} we will
mean Equation~\ref{list-def}.

Before we get to those interesting proofs, though, there are some other things
that need defining.
\AgdaDatatype{Fin}, for instance: \(\AgdaDatatype{Fin}\;n\) is the type of
natural numbers smaller than \(n\).
Its definition relies on a type for disjoint union,
\(\AgdaDatatype{\ensuremath{\uplus}}\).
\begin{multicols}{2} \null \vfill
  \begin{agdalisting}
    \ExecuteMetaData[agda/Data/Fin/Base.tex]{fin-def}
  \end{agdalisting} \vfill \null \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Introduction.tex]{disj-union}
  \end{agdalisting}%
  % It is also expressible with only \(\Sigma\):
  % \begin{agdalisting}
  %   \ExecuteMetaData[agda/Snippets/Introduction.tex]{sigma-disj-union}
  % \end{agdalisting}
  % Although the inductive type definition is slightly more ergonomic.
\end{multicols} \noindent
And containers themselves, of course.
Containers are a well-studied topic in dependent type theory, with a rich
theory: we won't dive in to that here.
\begin{definition}[Containers] \label{container-def}
  A container \cite{abbottContainersConstructingStrictly2005} is a pair
  \(S , P\) where \(S\) is a type, the elements of which are called
  the \emph{shapes} of the container, and \(P\) is a type family on \(S\), where
  the elements of \(P(s)\) are called the \emph{positions} of a container.
  We ``interpret'' a container into a functor defined like so:
  \begin{agdalisting} \label{container-interp}
    \ExecuteMetaData[agda/Container.tex]{container-interp}
  \end{agdalisting} \vspace{-\baselineskip}
\end{definition}

@@

As a slightly more complex example, consider the \(\AgdaDatatype{Fin}\) type
we've been using.
Remember that split enumerability is in fact the same as a split surjection from
\(\AgdaDatatype{Fin}\) (Lemma~\ref{split-enum-is-split-surj}): to show that
\(\AgdaDatatype{Fin}\) is split enumerable, then, we need only show that it has
a split surjection from itself.
We'll prove the following slightly more general statement:

\begin{agdalisting}
  \ExecuteMetaData[agda/Function/Surjective/Properties.tex]{split-surj-ident}
\end{agdalisting}
\subsection{Decidable Equality}
One thing that characterises all split enumerable types is that they are all
\emph{discrete}, i.e. they have decidable equality.  \todo{All multicol agda code should probably be vertically centred like this}
\begin{multicols}{2}  \null \vfill
  \begin{agdalisting}
    \ExecuteMetaData[agda/Relation/Nullary/Discrete/Base.tex]{discrete-def}
  \end{agdalisting} \vfill \null \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Dec.tex]{dec-def}
  \end{agdalisting}
\end{multicols} \vspace{-\baselineskip} \noindent
We will see later that this has implications for the space of types we're
dealing with, but for now it simply provides a useful function on split
enumerable types.
